ML Estimation of Matrix t Distributions with Expectation Maximization

### 2016-12-01 paper.tex
Minor adjustments.

\begin{abstract}
Matrix-$t$ distribution has applications in various areas of applied statistics, including recommendation systems, remote sensing, etc. In this article we provide an expectation-maximization (EM) formulae for Matrix-$t$ distributions and discuss their practical implementation.
\end{abstract}

\paragraph{Introduction}
Matrix-$t$ distributions have been applied to several predictive problems, such as spatial interpolation applied to the prediction of pollution concentration \cite{}, recommendation systems \cite{NIPS2007_3203},
...

@@

Most of the results are based on the integral

\begin{equation}
\int_{S\succ 0}e^{-\textrm{tr}SA}(\det S)^{x}dS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}\label{eq:intmultgammadef}
\end{equation}
where we integrate $S$ over the space of positive definite $p\times p$
matrices, $A$ is a $p\times p$ nonsingular matrix, and the multivariate
gamma function $\Gamma_{p}(a)$ is given in terms of the standard
gamma function by:
\begin{equation}
\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+(1-j)/2\right)\label{eq:multgammadef}
\end{equation}
we also define the multivariate digamma function
\[
\psi_{p}(a)=\frac{\partial}{\partial a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi(a+(1-j)/2)
\]

The PDF of a random matrix $p\times m$ $T$ following a matrix-t
distribution
\[
T_{p,m}(M,\Sigma,\Omega,n)
\]
is given by:
\begin{equation}
\begin{split}\textrm{tpdf}(T,M,\Sigma,\Omega,n)=\frac{\Gamma_{p}\left(\frac{p+n+m-1}{2}\right)}{\Gamma_{p}\left(\frac{n+p-1}{2}\right)\pi^{\frac{mp}{2}}(\det\Omega)^{\frac{p}{2}}(\det\Sigma)^{\frac{m}{2}}}\times\\
\det\left(I_{p\times p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}\right){}^{\frac{1-m-p-n}{2}}
\end{split}
\label{eq:matrixtpdf}
\end{equation}
where $\Sigma$ is a $p\times p$ positive definite matrix, $\Omega$
is a $m\times m$ positive definite matrix, $M$ is a $p\times m$
matrix, and $n$ is the number of degrees of freedom.

Following {[}book{]}, we begin by deriving the PDF of the matrix-t,
starting from a matrix normal distributed random matrix $X$, where
row-covariance matrix $S$ is distributed following a Wishart distribution
with the appropriate count of degrees of freedom (independent from
$X$), and marginalizing over $S$. We repeat the results from {[}book{]}
because, along the way, we will write several results that will be
useful in the derivation of the EM formulas.

The $p\times m$ random matrix $X$ follows the matrix-normal distribution

\begin{equation}
N_{p,m}(M,S,\Omega)\label{eq:ndist}
\end{equation}
if its PDF is given by:
\begin{equation}
\frac{1}{(2\pi)^{pm/2}(\det\Omega)^{p/2}(\det S)^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
\end{equation}
where $S$ is a $p\times p$ positive definite matrix (the covariance
of the rows of $X$), $\Omega$ is a $m\times m$ positive definite
matrix (the covariance of the columns of $X$) and $M$ is a $p\times m$
matrix (the mean of $X$).

In turn, the $p\times p$ matrix $S$ follows a Wishart distribution
\begin{equation}
W_{p}\left(\Sigma,\nu\right)\label{eq:wdist}
\end{equation}
 if its PDF is given by:
\begin{equation}
\frac{1}{2^{\nu p/2}(\det\Sigma)^{\nu/2}\Gamma_{p}(\frac{\nu}{2})}(\det S)^{\frac{\nu-p-1}{2}}\exp\left(-\textrm{tr}(\Sigma^{-1}S)\right)\label{eq:wdistpdf}
\end{equation}
where $\Sigma$ is a $p\times p$ positive definite matrix, $\nu>p-1$
is the degrees of freedom and $\Gamma_{p}$ is the multivariate gamma
function defined in \ref{eq:multgammadef}. We will show that the
random matrix
\begin{equation}
T=(S^{-\frac{1}{2}})^{t}X+M\label{eq:change}
\end{equation}
where $S^{\frac{1}{2}}$ is the Cholesky square root of $S$, follows
a matrix-$t$ distribution with $n=\nu-p+1$. The Jacobian of this
transformation is 
\begin{equation}
(\det S)^{\frac{1}{2}m}.\label{eq:jac}
\end{equation}

The joint PDF of $T$ and $S$ is the product of Eq.~\ref{eq:ndistpdf},
Eq.~\ref{eq:wdistpdf} and Eq.~\ref{eq:jac}:
\begin{equation}
\begin{split}j(T,S,M,\Sigma,\Omega,n)\coloneqq\frac{\pi^{-\frac{mp}{2}}(\det\Sigma)^{\frac{n+p-1}{2}}(\det\Omega)^{\frac{n+m-2}{2}}(\det S)^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
\exp\left(-\frac{1}{2}\textrm{tr}S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})\right)
\end{split}
\label{eq:jointpdf}
\end{equation}
Finally we obtain the PDF of the matrix-$t$ distribution Eq.~\ref{eq:matrixtpdf}
by integrating over $S$ in the above expression, using directly Eq.~\ref{eq:intmultgammadef}
and slightly rearranging the terms (in particular, noting that $\det(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})=\det\Sigma\det(I_{p\times p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}})$).

@@

We can split the integral in Eq.~\ref{eq:expdef} by the terms of
$\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating
a global factor $\frac{1}{2}$ and terms that do not depend on the
maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$,
we have: 
\[
\begin{split}\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\sim n_{t}\log\det S+\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))\\
-n_{t}p\log2-p\log\det\Omega_{t}+(n_{t}+p-1)\log\det\Sigma_{t}-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1))
\end{split}
\]
where the first two terms depend on $S$, and therefore we need to
compute the corresponding integral, while for the rest, the integral
is proportional to the PDF of the matrix-$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
In obtain to compute:
\begin{equation}
n_{t}\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log\det SdS\label{eq:firstlogpart}
\end{equation}
we note that taking the derivative with respect to $x$ in Eq.~\ref{eq:intmultgammadef},
we have:
\begin{equation}
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}\log\det SdS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}(\log\det A+\gamma_{p}(\frac{1}{2}(1+p+2x)))\label{eq:intderivative}
\end{equation}
We can solve \ref{eq:firstlogpart} by substituting this result directly.

>>>>

\begin{abstract}
Matrix--$t$ distribution has applications in various areas of applied statistics, including recommendation systems, remote sensing, etc. In this article we provide an expectation-maximization (EM) formulae for the maximum likelihood estimation of matrix--$t$ distributions and discuss their practical implementation.
\end{abstract}

\paragraph{Introduction}
Matrix--$t$ distributions have been applied to several predictive problems, such as spatial interpolation applied to the prediction of pollution concentration \cite{}, recommendation systems \cite{NIPS2007_3203}, and
...

@@

The derivation of EM formulae for matrix--$t$ distributions is mainly based on the following integral identity:

\begin{equation}
\int_{S\succ 0}e^{-\textrm{tr}(SA)}|S|^{x}dS=\Gamma_{p}\left(\frac{1+p+2x}{2}\right)|A|^{\frac{1+p+2x}{2}},\label{eq:intmultgammadef}
\end{equation}
where the domain of integration of $S$ is the set of real symmetric positive definite
matrices with dimension $p$, $A$ is a $p\times p$ nonsingular matrix, and $\Gamma_{p}(\cdot)$ is the multivariate gamma function given in terms of the standard gamma function $\Gamma(\cdot)$ by
\begin{equation}
\Gamma_{p}(a)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(a+\frac{1-j}{2}\right).\label{eq:multgammadef}
\end{equation}
For notation convenience, the matrix determinant is represented as $|\cdot|$. We also define the multivariate digamma function
\begin{equation}
\psi_{p}(a)=\frac{\partial}{\partial a}\log\Gamma_{p}(a)=\sum_{j=1}^{p}\psi\left(a+\frac{1-j}{2}\right).
\end{equation}

A random $p\times m$ matrix $T$ following a matrix--$t$ distribution $T_{p,m}(M,\Sigma,\Omega,n)$
has probability density
\begin{equation}
f(T;\left\{M,\Sigma,\Omega,n\right\})=\frac{\Gamma_{p}\left(\frac{p+n+m-1}{2}\right)}{\Gamma_{p}\left(\frac{n+p-1}{2}\right)\pi^{\frac{mp}{2}}|\Omega|^{\frac{p}{2}}|\Sigma|^{\frac{m}{2}}}
|I_p+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}}|^{\frac{1-m-p-n}{2}},
\label{eq:matrixtpdf}
\end{equation}
where $\Sigma$ and $\Omega$ are both real symmetric positive definite matrices of dimension $p$ and $m$, respectively, $M$ is a $p\times m$ matrix indicates the offset, and $n$ is the number of degrees of freedom.

Following {[}book{]} we begin by deriving the probability density of matrix--$t$ distribution starting from a matrix normal $N_{p,m}(M,S,\Omega)$ distributed random variate $X$ with Wishart $W_{p}\left(\Sigma,\nu\right)$ distributed row-scale parameter $S$ of appropriate degrees of freedom $\nu$ (independent from $X$), and marginalizing over $S$. Note that the row scale parameter $S$ is a real symmetric positive definite matrix. We repeat the results here because, along the way, we will utilize several results that are useful in the derivation of the EM formulas.

The $p\times m$ random matrix $X$ follows $N_{p,m}(M,S,\Omega)$ has probability density
\begin{equation}
\frac{1}{(2\pi)^{pm/2}(\det\Omega)^{p/2}(\det S)^{m/2}}\exp\left(-\frac{1}{2}\textrm{tr}\left[\Omega^{-1}(X-M)^{\textrm{t}}S^{-1}(X-M)\right]\right)\label{eq:ndistpdf}
\end{equation}
where $S$ is a $p\times p$ positive definite matrix (the covariance
of the rows of $X$), $\Omega$ is a $m\times m$ positive definite
matrix (the covariance of the columns of $X$) and $M$ is a $p\times m$
matrix (the mean of $X$).

In turn, the $p\times p$ matrix $S$ follows a Wishart distribution
\begin{equation}
W_{p}\left(\Sigma,\nu\right)\label{eq:wdist}
\end{equation}
 if its PDF is given by:
\begin{equation}
\frac{1}{2^{\nu p/2}(\det\Sigma)^{\nu/2}\Gamma_{p}(\frac{\nu}{2})}(\det S)^{\frac{\nu-p-1}{2}}\exp\left(-\textrm{tr}(\Sigma^{-1}S)\right)\label{eq:wdistpdf}
\end{equation}
where $\Sigma$ is a $p\times p$ positive definite matrix, $\nu>p-1$
is the degrees of freedom and $\Gamma_{p}$ is the multivariate gamma
function defined in \ref{eq:multgammadef}. We will show that the
random matrix
\begin{equation}
T=(S^{-\frac{1}{2}})^{t}X+M\label{eq:change}
\end{equation}
where $S^{\frac{1}{2}}$ is the Cholesky square root of $S$, follows
a matrix--$t$ distribution with $n=\nu-p+1$. The Jacobian of this
transformation is 
\begin{equation}
(\det S)^{\frac{1}{2}m}.\label{eq:jac}
\end{equation}

The joint PDF of $T$ and $S$ is the product of Eq.~\ref{eq:ndistpdf},
Eq.~\ref{eq:wdistpdf} and Eq.~\ref{eq:jac}:
\begin{equation}
\begin{split}j(T,S,M,\Sigma,\Omega,n)\coloneqq\frac{\pi^{-\frac{mp}{2}}(\det\Sigma)^{\frac{n+p-1}{2}}(\det\Omega)^{\frac{n+m-2}{2}}(\det S)^{\frac{m+n-2}{2}}}{2^{p\frac{n+m+p-1}{2}}\Gamma_{p}(\frac{n+p-1}{2})}\times\\
\exp\left(-\frac{1}{2}\textrm{tr}S(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})\right)
\end{split}
\label{eq:jointpdf}
\end{equation}
Finally we obtain the PDF of the matrix--$t$ distribution Eq.~\ref{eq:matrixtpdf}
by integrating over $S$ in the above expression, using directly Eq.~\ref{eq:intmultgammadef}
and slightly rearranging the terms (in particular, noting that $\det(\Sigma+(T-M)\Omega^{-1}(T-M)^{\textrm{t}})=\det\Sigma\det(I_{p\times p}+\Sigma^{-1}(T-M)\Omega^{-1}(T-M)^{\textrm{t}})$).

@@

We can split the integral in Eq.~\ref{eq:expdef} by the terms of
$\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})$. Eliminating
a global factor $\frac{1}{2}$ and terms that do not depend on the
maximization parameters $\{M_{t},\Sigma_{t},\Omega_{t},n_{t}\}$,
we have: 
\[
\begin{split}\log j(T_{i},S,M_{t},\Sigma_{t},\Omega_{t},n_{t})\sim n_{t}\log\det S+\textrm{tr}(S(\Sigma_{t}+(T_{i}-M)\Omega^{-1}(T_{i}-M)^{\textrm{t}}))\\
-n_{t}p\log2-p\log\det\Omega_{t}+(n_{t}+p-1)\log\det\Sigma_{t}-2\log\Gamma_{p}(\frac{1}{2}(n_{t}+p-1))
\end{split}
\]
where the first two terms depend on $S$, and therefore we need to
compute the corresponding integral, while for the rest, the integral
is proportional to the PDF of the matrix--$t$ distribution, Eq.~\ref{eq:matrixtpdf}.
In obtain to compute:
\begin{equation}
n_{t}\int_{S>0}j(T_{i},S,M,\Sigma,\Omega,n)\log\det SdS\label{eq:firstlogpart}
\end{equation}
we note that taking the derivative with respect to $x$ in Eq.~\ref{eq:intmultgammadef},
we have:
\begin{equation}
\int_{S>0}e^{-\textrm{tr}SA}(\det S)^{x}\log\det SdS=\Gamma_{p}\left(\frac{1}{2}(1+p+2x)\right)(\det A)^{\frac{1}{2}(1+p+2x)}(\log\det A+\gamma_{p}(\frac{1}{2}(1+p+2x)))\label{eq:intderivative}
\end{equation}
We can solve \ref{eq:firstlogpart} by substituting this result directly.
